Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-07-01 PandaEvolution.tex
introduce  \ifreview
  to toggle between reviewer comments/not, so as to make it easier to get a handle on
  net space used.
Address referees' comments and further reductions

The upcoming LHC Run 3 will require more resources than the Worldwide LHC
Computing Grid (WLCG) can provide. Currently, PanDA WMS uses more than 300,000
cores at over 100 Grid sites, with a peak performance of 0.3 petaFLOPS. This
capacity will be sufficient for the planned analysis and data processing, but it
will be insufficient for the Monte Carlo production workflow and any extra
activity. To alleviate these challenges, ATLAS is expanding the current
computing model to include additional resources such as the opportunistic use of
supercomputers.

Generally, supercomputers are designed to support parallel computation that
requires runtime communication. Jobs are executed across multiple cores, each
core calculating a small part of the problem and communicating with other cores
via MPI. Accordingly, supercomputers have large number of worker nodes,
connected through a high-speed, low-latency dedicated network. Each worker node
has multicore CPUs, usually augmented with Graphics Processing Units (GPUs) or
other specialized coprocessors.

@@

We developed and deployed a single-point solution to better understand the
problem space of enabling a WMS designed for HTC to execute production workflows
on resources designed to support HPC. The PanDA team developed a job broker to
support the execution of part of the ATLAS production Monte Carlo workflow on
Titan, a leadership-class supercomputer managed by the Oak Ridge Leadership
Computing Facility (OLCF) at the Oak Ridge National Laboratory (ORNL).

% -----------------------------------------------------------------------------
\subsection{Architectures and Interfaces}
\label{ssec:panda-titan}

The Titan supercomputer, current number three on the Top 500 list~\cite{top500},
is a Cray XK7 system with 18,688 worker nodes and a total of 299,008 CPU cores.
Each worker node has an AMD Opteron  6274 16-core CPU, a Nvidia Tesla K20X GPU,
32 GB of RAM and no local storage, though a 16 GB RAM disk can be set up. Work
nodes use Cray’s Gemini interconnect for inter-node MPI messaging. Titan is
served by the Spider II~\cite{oral2013olcf}, a Lustre filesystem with 32 PB of
disk storage, and by a 29 PB HPSS tape storage system. Titan’s worker nodes run
Compute Node Linux, a run time environment based on SUSE Linux Enterprise
Server.

Titan's users submit jobs to Titan's PBS scheduler by logging into login or data
transfer nodes (DTNs). Titan's authentication and authorization model is based
on two-factor authentication with a RSA SecurID key. Login nodes and DTNs have
out/inbound wide area network connectivity while worker nodes have only local
network access. Fair-share and allocation policies are in place both for the PBS
batch system and shared file systems.

Titan's architecture, configuration and policies poses several challenges to the
integration with PanDA. The default deployment
model of PanDA Pilot is unfeasible on Titan: PanDA Pilot is required to contact
the Job Dispatcher of the PanDA Server to pull jobs to execute, but this is not
possible on Titan because worker nodes do not offer outbound network
connectivity. Further, Titan does not support PanDA's security model based on
certificates and virtual organizations, making PanDA's approach to identity
management also unfeasible. While DTNs offer wide area network data transfer, an
integration with ATLAS DDM is beyond the functional and administrative scope of
the current prototyping phase. Finally, the specific characteristics of the
execution environment, especially the absence of local storage on the worker
nodes and modules tailored to Compute Node Linux, require re-engineering of
ATLAS application frameworks.

Currently, very few HEP applications can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks of ATLAS workflows can
be off-loaded from the Grid to Titan's. Further, when HEP tasks can be
partitioned into independent jobs, Titan worker nodes can be used to execute up
to 16 concurrent payloads, one per each available core. Given these constraints
and challenges, the type of task most suitable for execution at the moment on
Titan is Monte Carlo detector simulation. This type of task is mostly
computational-intensive, requiring less than 2GB of RAM at runtime and with
small input data requirements. Detector simulation tasks in ATLAS are performed
via AthenaMP~\cite{aad2010atlas}, the ATLAS software framework integrating the
GEANT4 simulation toolkit~\cite{agostinelli2003geant4}. These tasks account for
$\approx$ 60\% of all the jobs on WLCG, making them a primary candidate for
offloading.

Detector simulation is part of the ATLAS production Monte Carlo (MC)
workflow~\cite{rimoldi2006atlas,de2013delphes,ritsch2014atlas}. The MC workflow
consists of four main stages: event generation, detector simulation,
digitization, and reconstruction. Event generation creates sets of particle
four-momenta via different generators, e.g., PYTHIA~\cite{sjostrand2006pythia},
HERWIG~\cite{corcella2001herwig} and many others. The detector simulator is
called Geant4~\cite{agostinelli2003geant4} and simulates the ATLAS detector and
the interaction between that detector and particles. Each interaction creates a
so-called hit and all hits are collected and passed on for digitalization, where
hits are further processed to mimic the readout of the detector. Finally,
reconstruction operates local pattern recognition, creating high-level objects
like particles and jets.

% -----------------------------------------------------------------------------
\subsection{PanDA Broker}
\label{ssec:panda_titan}

@@

The backfill optimization of Titan's Moab scheduler allows to avoid the overhead
of queue wait times without using pilot abstraction~\cite{maui_backfill_url}.
With this optimization, Moab starts low-priority jobs when they do not delay
higher priority jobs, independent of whether the low-priority jobs were queued
after the high-priority jobs.

@@

At any point in time, the size of an available backfill slot is typically a
small fraction of the total capacity of a resource. Notwithstanding, given the
size of Titan this translates into a substantial capacity. Every year, about
10\% of Titan's capacity remains unused~\cite{barker2016us}, corresponding to an
average of 30,000 unused cores (excluding GPU cores). This equals to roughly
10\% of the overall capacity of WLCG.

Given the communication requirements of PanDA Pilots and the unused capacity of
Titan, PanDA pilot was repurposed to serve as a job broker on the DTN nodes of
Titan (Fig.~\ref{fig:panda_broker}). Maintaining the core modules of PanDA Pilot
and its stand-alone architecture, this prototype called `PanDA Broker'
implements functionalities to: (i) interrogate Titan about backfill
availability; (ii) pull ATLAS jobs and events from PanDA Server; (iii) wrap the
payload of ATLAS jobs into MPI scripts; (iv) submitting MPI scripts to Titan's
PBS batch system and monitor their execution; and (v) staging and preparing
input/output files. Backfill querying, payload wrapping, and scripts submission
required a new implementation while pulling ATLAS job and events, and file
staging were inherited from PanDA Pilot.

Backfill querying is performed via a dedicated Moab scheduler command while a
tailored Python MPI script is used to execute the payload of ATLAS jobs. This
MPI script enables the execution of unmodified Grid-centric, ATLAS jobs on
Titan. Typically, a MPI script is workload-specific as it sets up the execution
environment for a specific payload. This involves organization of worker
directories, data management, optional input parameters modification, and
cleanup on exit. Upon submission, a copy of the MPI script runs on every
available worker node, starting the execution of the ATLAS job's payload in a
subprocess and waits until its completion.

MPI scripts are submitted to Titan's PBS batch system via
RADICAL-SAGA~\cite{radical-saga_url}, a Python module, compliant with the OGF
GFD.90 SAGA specification~\cite{goodale2008simple}. The Simple API for Grid
Applications (SAGA) offers a unified interface to diverse job schedulers and
file transferring services. In this way, SAGA provides an interoperability layer
that lowers the complexity of using distributed infrastructures. Behind the API
façade, RADICAL-SAGA implements a adaptor architecture: each adaptor interface
the SAGA API with different middleware systems and services, including the PBS
batch scheduler of Titan.

The data staging capabilities of the PanDA Broker are implemented via a file
system that is shared among DTNs and worker nodes. The input files with the
events of the ATLAS jobs are downloaded on the shared filesystem from the data
center of Brookhaven National Laboratory (BNL). The MPI script setup process
includes making the location of these files available to the payload of the
ATLAS's jobs. The PanDA Broker can locate the payload's output files on the
shared filesystem and transfer them from Titan BNL.

Once deployed on Titan, every PanDA Broker supports the execution of MC detector
simulations in 9 steps. PanDA Broker queries the Job Dispatcher module of the
PanDA server for ATLAS jobs that have been bound to Titan by JEDI
(Fig.~\ref{fig:panda_broker}:1). Upon receiving jobs descriptions, PanDA Broker
pulls jobs' input files from BNL to the OLCF Lustre file system
(Fig.~\ref{fig:panda_broker}:2). PanDA Broker queries Titan's Moab scheduler
about the current available backfill slot (Fig.~\ref{fig:panda_broker}:3) and
creates an MPI script, wrapping enough ATLAS jobs' payload to fit the backfill
slot. PanDA Broker submits the MPI script to the Titan's Torque batch system via
RADICAL-SAGA (Fig.~\ref{fig:panda_broker}:4).

>>>>

% \ifreview
% Comments to address in this section:
% \begin{enumerate}
% 	\item REVIEWER 2: The authors spent quite a few paragraphs for the detail
% 	design and implementation of the PanDA WMS, but introduced few background
% 	of the LHC applications. According to the paper, I understand the
% 	applications running on Titan are a large amount of small independent
% 	jobs. Thus, it seems less cost-effective to run these jobs on a
% 	world-leading supercomputer like Titan which has a fast and expensive
% 	interconnection. It will be more interesting if the authors can compare
% 	the solution on Titan (proposed in this paper) with the original
% 	distributed computing environment and HPECloud from Google mentioned in
% 	the related work section.
% 	\item REVIEWER 5: In Section 3.2, GPUs are not considered in this work,
% 	which would be a waste for using HPC computing. Could this work be
% 	extended to include GPUs or other heterogenous resources?
% \end{enumerate}
% \fi
The upcoming LHC Run 3 will require more resources than the Worldwide LHC
Computing Grid (WLCG) can provide. Currently, PanDA WMS uses more than
300,000 cores at over 100 Grid sites, with a peak performance of 0.3
petaFLOPS\@. This capacity will be sufficient for the planned analysis and
data processing, but it will be insufficient for the Monte Carlo production
workflow and any extra activity. To alleviate these challenges, ATLAS is
expanding the current computing model to include additional resources such as
the opportunistic use of supercomputers.

Generally, supercomputers are designed to support parallel computation that
requires runtime communication. Jobs are executed across multiple cores, each
core calculating a small part of the problem and communicating with other
cores via MPI\@. Accordingly, supercomputers have large number of worker
nodes, connected through a high-speed, low-latency dedicated network. Each
worker node has multicore CPUs, usually augmented with Graphics Processing
Units (GPUs) or other specialized coprocessors.

@@

Currently, ATLAS workloads do not require fast interconnects or specilized
coprocessors but supercomputers tend not to reach 100\% utilization due to
the scheduling of jobs requiring large amount of resources. This offers the
possibility to execute ATLAS-like workloads on supercomputers to increase
utilization and reducing the waste of available resources.

We developed a single-point solution to better understand the problem space
of enabling a WMS designed for HTC to execute production workflows on
resources designed to support HPC\@. The PanDA team developed a job broker to
support the execution of part of the ATLAS production Monte Carlo workflow on
Titan, a leadership-class supercomputer managed by the Oak Ridge Leadership
Computing Facility (OLCF) at the Oak Ridge National Laboratory (ORNL).

% ---------------------------------------------------------------------------
\subsection{Architectures, Interfaces and Workloads}
\label{ssec:panda-titan}

% The Titan supercomputer, current number three on the Top 500
% list,%~\cite{top500}, 
% is a Cray XK7 system with 18,688 worker nodes and a total
% of 299,008 AMD cores and 18,688 Nvidia Tesla GPUs. 
% Each worker node has an AMD Opteron 6274 16-core CPU,
% a Nvidia Tesla K20X GPU, 32 GB of RAM and no local storage, though a 16 GB
% RAM disk can be set up. Work nodes use Cray’s Gemini interconnect for
% inter-node MPI messaging. Titan is served by the Spider
% II~\cite{oral2013olcf}, a Lustre filesystem with 32 PB of disk storage, and
% by a 29 PB HPSS tape storage system. Titan’s worker nodes run Compute Node
% Linux, a run time environment based on SUSE Linux Enterprise Server.

% Titan's users submit jobs to Titan's PBS scheduler by logging into login or
% data transfer nodes (DTNs). Titan's authentication and authorization model
% is based on two-factor authentication with a RSA SecurID key. Login nodes
% and DTNs have out/inbound wide area network connectivity while worker nodes
% have only local network access. Fair-share and allocation policies are in
% place both for the PBS batch system and shared file systems.

Titan's architecture, configuration and policies poses several challenges to
the integration with PanDA\@. The default deployment model of PanDA Pilot is
unfeasible on Titan: PanDA Pilot is required to contact the Job Dispatcher of
the PanDA Server to pull jobs to execute, but this is not possible on Titan
because worker nodes do not offer outbound network connectivity. Further,
Titan does not support PanDA's security model based on certificates and
virtual organizations, making PanDA's approach to identity management also
unfeasible. While Titan's data transfer nodes (DTNs) offer wide area network
data transfer, an integration with ATLAS DDM is beyond the functional and
administrative scope of the current prototyping phase. Finally, the specific
characteristics of the execution environment, especially the absence of local
storage on the worker nodes and modules tailored to Compute Node Linux,
require re-engineering of ATLAS application frameworks.

Currently, very few HEP applications can benefit from Titan's GPUs but some
computationally-intensive and non memory-intensive tasks of ATLAS workflows
can be off-loaded from the Grid to Titan's. Further, when HEP tasks can be
partitioned into independent jobs, Titan worker nodes can be used to execute
up to 16 concurrent payloads, one per each available core. Given these
constraints and challenges, the type of task most suitable for execution at
the moment on Titan is Monte Carlo detector simulation. This type of task is
mostly computational-intensive, requiring less than 2GB of RAM at runtime and
small input data. Detector simulation tasks in ATLAS are performed via
AthenaMP~\cite{aad2010atlas}, the ATLAS software framework integrating the
GEANT4 detector simulation toolkit~\cite{agostinelli2003geant4}. These tasks
account for $\approx$ 60\% of all the jobs on WLCG, making them a primary
candidate for offloading.

Detector simulation is part of the ATLAS production Monte Carlo (MC)
workflow~\cite{rimoldi2006atlas}.%,de2013delphes,ritsch2014atlas}. 
The MC workflow consists of four main stages: event generation, detector
simulation, digitization, and reconstruction. Event generation creates sets
of particle four-momenta via different generators, e.g.,
PYTHIA,%~\cite{sjostrand2006pythia}, 
HERWIG,%~\cite{corcella2001herwig} 
and many others. Geant4 simulates the ATLAS detector and the interaction
between that detector and particles. Each interaction creates a so-called hit
and all hits are collected and passed on for digitalization, where hits are
further processed to mimic the readout of the detector. Finally,
reconstruction operates local pattern recognition, creating high-level
objects like particles and jets.

% ---------------------------------------------------------------------------
\subsection{PanDA Broker}
\label{ssec:panda_titan}

@@

The backfill optimization of Titan's Moab scheduler allows to avoid the
overhead of queue wait times without using pilot
abstraction~\cite{maui_backfill_url}. With this optimization, Moab starts
low-priority jobs when they do not delay higher priority jobs, independent of
whether the low-priority jobs were queued after the high-priority ones.

@@

At any point in time, the size of an available backfill slot is typically a
small fraction of the total capacity of a resource. Notwithstanding, given
the size of Titan this translates into a substantial capacity. Every year,
about 10\% of Titan's capacity remains unused~\cite{barker2016us},
corresponding to an average of 30,000 unused cores (excluding GPU cores).
This equals to roughly 10\% of the overall capacity of WLCG\@.

Given the communication requirements of PanDA Pilots and the unused capacity
of Titan, PanDA pilot was repurposed to serve as a job broker on the DTN
nodes of Titan (Fig.~\ref{fig:panda_broker}). Maintaining the core modules of
PanDA Pilot and its stand-alone architecture, this prototype called `PanDA
Broker' 
% implements functionalities to: (i) interrogate Titan about backfill
% availability via a dedicated Moab scheduler command; (ii) pull ATLAS jobs
% and events from PanDA Server; (iii) wrap the payload of ATLAS jobs into MPI
% scripts; (iv) submitting workload-specific MPI scripts to Titan's PBS batch
% system and monitor their execution; and (v) staging and preparing
% input/output files. Backfill querying, payload wrapping, and scripts
% submission required a new implementation while pulling ATLAS job and
% events, and file staging were inherited from PanDA Pilot.
%
% Backfill querying is performed via a dedicated Moab scheduler command while
% a tailored Python MPI script is used to execute the payload of ATLAS jobs.
% This enables the execution of unmodified Grid-centric, ATLAS jobs on Titan.
% Typically, a MPI script is workload-specific as it sets up the execution
% environment for a specific payload. This involves organization of worker
% directories, data management, optional input parameters modification, and
% cleanup on exit. Upon submission, a copy of the MPI script runs on every
% available worker node, starting the execution of the ATLAS job's payload in
% a subprocess and waits until its completion.
%
% MPI scripts are submitted to Titan's PBS batch system via
% RADICAL-SAGA~\cite{radical-saga_url}, a Python module, compliant with the
% OGF GFD.90 SAGA specification~\cite{goodale2008simple}. The Simple API for
% Grid Applications (SAGA) offers a unified interface to diverse job
% schedulers and file transferring services. In this way, SAGA provides an
% interoperability layer that lowers the complexity of using distributed
% infrastructures. Behind the API façade, RADICAL-SAGA implements a adaptor
% architecture: each adaptor interface the SAGA API with different middleware
% systems and services, including the PBS batch scheduler of Titan.
%
% The data staging capabilities of the PanDA Broker are implemented via a
% file system shared among DTNs and worker nodes. The input files with the
% events of the ATLAS jobs are downloaded on the shared filesystem from the
% data center of Brookhaven National Laboratory (BNL). The MPI script setup
% process includes making the location of these files available to the
% payload of the ATLAS's jobs. The PanDA Broker can locate the payload's
% output files on the shared filesystem and transfer them from Titan BNL\@.
%
% Once deployed on Titan, every PanDA Broker 
supports the execution of MC detector simulations in 9 steps.

Each PanDA Broker deployed on Titan queries the Job Dispatcher module of the
PanDA server for ATLAS jobs that have been bound to Titan by JEDI
(Fig.~\ref{fig:panda_broker}:1). Upon receiving jobs descriptions, PanDA
Broker pulls jobs' input files from BNL to the OLCF Lustre file system
(Fig.~\ref{fig:panda_broker}:2). PanDA Broker queries Titan's Moab scheduler
about the current available backfill slot (Fig.~\ref{fig:panda_broker}:3) and
creates an MPI script, wrapping enough ATLAS jobs' payload to fit the
backfill slot. PanDA Broker submits the MPI script to the Titan's Torque
batch system via RADICAL-SAGA (Fig.~\ref{fig:panda_broker}:4).
